Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-22 ms.tex
small typo corrected.
Frank's final changes incorporated, including removing Prince from the author's list.

\author[Zentner et al.]
{Andrew R. Zentner$^{1}$, Andrew Hearin$^{2}$, Frank C. van den Bosch$^{3},$ \newauthor
Johannes U. Lange$^{3}$, Antonio Villarreal$^{1}$, and P. Rogers Nelson$^{4}$\\ \\
$^1$Department of Physics and Astronomy \& Pittsburgh Particle Physics, Astrophysics, and Cosmology Center (PITT PACC),\\ University of Pittsburgh, Pittsburgh, PA 15260\\
$^2$Yale Center for Astronomy \& Astrophysics, Yale University, New Haven, CT\\
$^3$Department of Astronomy, Yale University, P.O. Box 208101, New Haven, CT\\
$^4$Paisley Park, Chanhassen, MN\\
}

@@

This work presents a re-analysis of the SDSS DR7 data that
overcomes numerous shortcomings of previous studies.
The primary improvements of our approach stem from the fact that we 
directly populate mock galaxy catalogs within a cosmological simulation,
whereas previous analyses relied upon analytical fitting functions
with parameters calibrated against a suite of simulations. 
In direct mock population, delicate issues present in analytic
modeling, such as scale-dependent halo bias and halo exclusion, 
are treated exactly. 
Our approach also provides a natural framework for studying 
models of assembly bias, because we can use the exact clustering 
properties of simulated halos as a function of
any arbitrary halo properties in the catalog. It is not necessary to develop 
new phenomenological models or fitting functions to deal with this additional 
complexity. With direct mock population, systematic uncertainty in the 
model is limited to the sample variance of the simulated box,
numerical inaccuracies of halo-finding, and errors related 
to insufficient numerical resolution \citep[e.g., overmerging; ][]{klypin99a,rockstar}.

@@

The results of this subsection demonstrate that we achieve reasonable fits to projected galaxy clustering
data using direct HOD population of a high-resolution numerical simulation of structure formation. These results
also update and supersede existing constraints in the literature in at least three respects. First, we work within the
best-fit Planck cosmology. Second, we perform our parameter inference analysis using direct population of halos
identified in a numerical simulation of cosmological structure formation (BolshoiP). This greatly mitigates modeling
uncertainties associated with nonlinear density field evolution, scale-dependent halo bias, halo exclusion,
or other effects that have been difficult to incorporate into analytical halo models with high precision
\citep[e.g.,][and references therein]{vdBosch13}. Third,
we have explored the posteriors of the parameters with significantly more samples (roughly two orders of magnitude),
thereby mitigating errors on inferred parameters caused by insufficient sampling of the
posterior by \citet{zehavi_etal11}.

@@

%------------------------------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr19ABTri.pdf}
\caption{ Two-dimensional marginalized constraints on decorated HOD
  parameters inferred from fits to $\wprp(\rp)$ data for the $M_r<-19$
  sample. The contours and histograms along the diagonal panels are as
  in Fig.~\ref{fig:Mr20triangle}. The dashed line in the panels along the diagonal
  show the posteriors on these parameters from the standard analysis without
  any parameterization of galaxy assembly bias. In these panels, both histograms 
  are normalized to the same total probability. The decorated HOD models include a
  two-parameter model for assembly bias. The HOD parameter 
  $\log (M_0)$ is extremely poorly constrained by the data and has been
  suppressed for clarity. Likewise, as in Fig.~\ref{fig:Mr20triangle},
  $\sigma_{\log M}$ and $\log (M_{\rm min})$ share a narrow
  degeneracy, so we have suppressed $\sigma_{\log M}$ in order to make
  constraints on other parameters more easily visible.  }
\label{fig:Mr19ABtriangle}
\end{center}
\end{figure*}
%----------------------------------------------------------------------------------------------

@@

Beyond those generic conclusions, a few specific cases are worthy of further examination.
Consider the $M_r < -20$ sample. The inferred value of $A_{\rm cen} > 0.28$ at 99\%
confidence. In this case the data strongly prefer $A_{\rm cen} > 0$ ($A_{\rm cen} > 0$ at $99.9\%$) 
and thus strongly prefer galaxies to reside in halos of larger concentration at fixed halo mass. 
Figure~\ref{fig:directhodcomparison}
compares the inferred HODs for the $M_r<-20$ sample in a standard HOD analysis with the
inferred HOD from our decorated HOD analysis that includes parameters to describe
assembly bias. The bands in purple rein in the region of HODs for models within 
$\Delta chi^2 \le 5.89$ of the best-fit model. 
This is an explicit example of the degree to which the inferred relationship
between galaxies and halos can be altered by assembly bias. In the assembly-biased case, the efficiency 
of forming a sufficiently luminous galaxy is a function that increases slowly over nearly two 
decades in halo mass in contrast to a standard analysis HOD where this efficiency rises 
rapidly over roughly one third of a decade in $\mvir$. Taking these results at face value, 
the reason that the HOD of central galaxies can vary slowly over several orders of magnitude 
in $\mvir$ is that larger galaxies form more efficiently in high-concentration halos at fixed 
mass.

@@

Our most important conclusions are as follows.
%
\ben
%
\item It is not possible to rule out galaxy assembly bias using SDSS DR7 measurements of $\wprp$ and $n_{\rm g}.$
%
\item Decorated HOD fits to $\wprp$ and $n_{\rm g}$ favor significant levels of assembly bias, particularly in the lower luminosity thresholds we study. Both the $\magr<-20, -20.5$ samples prefer relatively strong central galaxy assembly bias. The $\magr<-19$ sample favors satellite assembly bias. Assembly bias in these samples improves the capability of the model to describe galaxy clustering near
the one-halo to two-halo transition at $r_{\rm p} \sim 2 \, h^{-1}\mathrm{Mpc}$.
%
\item The evidence for galaxy assembly bias weakens for brighter galaxy samples: 
Decorated HOD fits to the $\magr<-21$ sample are consistent with both $A_{\rm cen}=0$ and $A_{\rm sats}=0.$ 
This is consistent with the well-established result that
{\em halo assembly bias} weakens with increasing halo mass over the dynamic range relevant to these galaxy samples
\citep[see, e.g., Figure 8 of][and references therein]{hearin_etal16}.
%
\item Our posteriors and best-fit parameters summarized in
Table \ref{table:parameters} supersede the values published in \citet{zehavi_etal11},
as direct-mock-population together with the Decorated HOD allows us to account for
highly significant systematics that have heretofore been neglected from all HOD fits to SDSS data.
We note that our findings update the original \citet{zehavi_etal11} results
{\em even for our fits to standard HODs without assembly bias}, because
the original results were derived from MCMC chains that did not sufficiently sample the HOD model posteriors.
%
\een
%

>>>>

\author[Zentner et al.]
{Andrew R. Zentner$^{1}$, Andrew Hearin$^{2}$, Frank C. van den Bosch$^{3},$ \newauthor
Johannes U. Lange$^{3}$, and Antonio Villarreal$^{1}$\\ \\
$^1$Department of Physics and Astronomy \& Pittsburgh Particle Physics, Astrophysics, and Cosmology Center (PITT PACC),\\ University of Pittsburgh, Pittsburgh, PA 15260\\
$^2$Yale Center for Astronomy \& Astrophysics, Yale University, New Haven, CT\\
$^3$Department of Astronomy, Yale University, P.O. Box 208101, New Haven, CT\\
}

@@

This work presents a re-analysis of the SDSS DR7 data that
overcomes numerous shortcomings of previous studies.
The primary improvements of our approach stem from the fact that we 
populate directly halos within a cosmological simulation with mock galaxies, 
whereas previous analyses relied upon analytical fitting functions
with parameters calibrated against a suite of simulations. 
In direct mock population, delicate issues present in analytic
modeling, such as scale-dependent halo bias and halo exclusion, 
are treated exactly. 
Our approach also provides a natural framework for studying 
models of assembly bias, because we can use the exact clustering 
properties of simulated halos as a function of
any arbitrary halo properties in the catalog. It is not necessary to develop 
new phenomenological models or fitting functions to deal with this additional 
complexity. With direct mock population, systematic uncertainty in the 
model is limited to the sample variance of the simulated box,
numerical inaccuracies of halo-finding, and errors related 
to insufficient numerical resolution \citep[e.g., overmerging; ][]{klypin99a,rockstar}.

@@

The results of this subsection demonstrate that we achieve reasonable fits to projected galaxy clustering
data using direct HOD population of a high-resolution numerical simulation of structure formation. These results
also update and supersede existing constraints in the literature in at least three respects. First, we work within the
best-fit Planck cosmology. Second, we perform our parameter inference analysis using direct population of halos
identified in a numerical simulation of cosmological structure formation (BolshoiP). This greatly mitigates modeling
uncertainties associated with nonlinear density field evolution, scale-dependent halo bias, halo exclusion,
or other effects that have been difficult to incorporate into analytical halo models with high precision
\citep[e.g.,][and references the]{vdBosch13}. Third,
we have explored the posteriors of the parameters with significantly more samples (roughly two orders of magnitude),
thereby mitigating errors on inferred parameters caused by insufficient sampling of the
posterior by \citet{zehavi_etal11}.

@@

%------------------------------------------------------------------------------------------------
\begin{figure*}
\begin{center}
\includegraphics[width=15.0cm]{Mr19ABTri.pdf}
\caption{ Two-dimensional marginalized constraints on decorated HOD
  parameters inferred from fits to $\wprp(\rp)$ data for the $M_r<-19$
  sample. The contours and histograms along the diagonal panels are as
  in Fig.~\ref{fig:Mr20triangle}. The dashed line in the panels along the diagonal
  show the posteriors on these parameters from the standard analysis without
  any parameterization of galaxy assembly bias. In these panels, both histograms 
  are normalized to the same total probability. The decorated HOD models include a
  two-parameter model for assembly bias. The HOD parameter 
  $\log (M_0)$ is extremely poorly constrained by the data and we omit it for 
  clarity. Likewise, as in Fig.~\ref{fig:Mr20triangle},
  $\sigma_{\log M}$ and $\log (M_{\rm min})$ share a narrow
  degeneracy, so we have suppressed $\sigma_{\log M}$ in order to make
  constraints on other parameters more easily visible.  }
\label{fig:Mr19ABtriangle}
\end{center}
\end{figure*}
%----------------------------------------------------------------------------------------------

@@

Beyond those generic conclusions, a few specific cases are worthy of further examination.
Consider the $M_r < -20$ sample. The inferred value of $A_{\rm cen} > 0.28$ at 99\%
confidence. In this case the data strongly prefer $A_{\rm cen} > 0$ ($A_{\rm cen} > 0$ at $99.9\%$) 
and thus strongly prefer galaxies to reside in halos of larger concentration at fixed halo mass. 
Figure~\ref{fig:directhodcomparison}
compares the inferred HODs for the $M_r<-20$ sample in a standard HOD analysis with the
inferred HOD from our decorated HOD analysis that includes parameters to describe
assembly bias. The bands in purple rein in the region of HODs for models within 
$\Delta \chi^2 \le 5.89$ of the best-fit model. 
This is an explicit example of the degree to which the inferred relationship
between galaxies and halos can be altered by assembly bias. In the assembly-biased case, the efficiency 
of forming a sufficiently luminous galaxy is a function that increases slowly over nearly two 
decades in halo mass in contrast to a standard analysis HOD where this efficiency rises 
rapidly over roughly one third of a decade in $\mvir$. Taking these results at face value, 
the reason that the HOD of central galaxies can vary slowly over several orders of magnitude 
in $\mvir$ is that larger galaxies form more efficiently in high-concentration halos at fixed 
mass.

@@

Our most important conclusions are as follows.
%
\ben
%
\item It is not possible to rule out galaxy assembly bias using SDSS DR7 measurements of $\wprp$ and $n_{\rm g}$.
%
\item Decorated HOD fits to $\wprp$ and $n_{\rm g}$ favor significant levels of assembly bias, particularly in the lower luminosity thresholds we study. Both the $\magr<-20, -20.5$ samples prefer relatively strong central galaxy assembly bias. The $\magr<-19$ sample favors satellite assembly bias. Assembly bias in these samples improves the capability of the model to describe galaxy clustering near
the one-halo to two-halo transition at $r_{\rm p} \sim 2 \, h^{-1}\mathrm{Mpc}$.
%
\item The evidence for galaxy assembly bias weakens for brighter galaxy samples: 
Decorated HOD fits to the $\magr<-21$ sample are consistent with both $A_{\rm cen}=0$ and $A_{\rm sat}=0.$ 
This is consistent with the well-established result that
{\em halo assembly bias} weakens with increasing halo mass over the dynamic range relevant to these galaxy samples
\citep[see, e.g., Figure 8 of][and references therein]{hearin_etal16}.
%
\item Our posteriors and best-fit parameters summarized in
Table \ref{table:parameters} supersede the values published in \citet{zehavi_etal11},
as direct-mock-population together with the Decorated HOD allows us to account for
highly significant systematics that have heretofore been neglected from all HOD fits to SDSS data.
We note that our findings update the original \citet{zehavi_etal11} results
{\em even for our fits to standard HODs without assembly bias}, because
the original results were derived from MCMC chains that did not sufficiently sample the HOD model posteriors.
%
\een
%
